GPU-Beschleunigung der Material Point Method

### 2018-11-27 mscthesis.tex
hessian

@@

  h(\boldsymbol{\hat{v}}_i^{n+1}) = m_i^n \frac{\boldsymbol{\hat{v}_i}^{n+1}-\boldsymbol{v}_i^n}{\Delta t} - \boldsymbol{f}_i\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{\hat{v}}_i^{n+1} + \boldsymbol{v}_i^{n})\right) = 0.

@@

In general starting off with an energy description is a more physical approach of the problem. The first term can be identified clearly as the kinetic energy. Moreover it easily allows adding potential terms for gravity or collision \cite{MPM:OPTIMI_INTEGR}. Or modifying the energy function altogether based off for instance the fundamental state of matter: as an example the phase transition to and physics of a liquid \cite{MPM:PHASE_CHANGE}.
The analysis of minimizing the objective is equivalent to finding the zero crossing of the derivative:
\begin{equation}
  \argmin_{\forall \boldsymbol{v}_i}(E(\boldsymbol{v}_i)) \Leftrightarrow g(\boldsymbol{v}_i) = \frac{\partial E}{\partial \boldsymbol{v}_i} = 0.
\end{equation}
Since the objective from the start was minimizing $\boldsymbol{E}$ to find $\boldsymbol{v}_i$ the scale of this function can be chosen arbitrarily, i.e. a zero crossing does not scale. The scaling here can be identified as:
\begin{equation}
  g(\boldsymbol{v}_i) = \frac{\Delta t}{4} h(\boldsymbol{v}_i) = m_i^n \frac{\boldsymbol{v_i}-\boldsymbol{v}_i^n}{4} -\frac{\Delta t}{4} \boldsymbol{f}_i\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}_i + \boldsymbol{v}_i^{n})\right).
\end{equation}
\cite{MPM:COURSE}\cite{MPM:APIC}

\subsection{GPU memory}
\subsubsection{Optimization opportunities}
\cite{NVIDIA:BEST:PRACTICE}
\cite{AMD:GPU_OPEN}
\begin{itemize}
  \item Minimize CPU-GPU transfers
  \item Coalesced memory access
  \item Bank conflicts
\end{itemize}
\subsubsection{SoA vs. AoS}
\section{Implementation}
\newcommand{\var}[1]{{\ttfamily#1}}% variable
\begin{algorithm}[t]
  \caption{Conjugate gradient}\label{alg:conj_grad}
  \begin{algorithmic}[1]
    \Procedure{CONJUGATE-GRADIENT}{$A,x,b$}
    \State $\boldsymbol{r}\gets \boldsymbol{b} - \boldsymbol{Ax}$
    \State $\boldsymbol{p}\gets \boldsymbol{r}$
    \State $\gamma \gets \langle \boldsymbol { r } , \boldsymbol { r } \rangle$
    \Repeat
      \State $\boldsymbol { s } \gets \boldsymbol { A } \boldsymbol { p }$
      \State $\alpha \gets \frac { \gamma } { \langle \boldsymbol { p } , \boldsymbol { s } \rangle }$
      \State $\boldsymbol { x } \gets \boldsymbol { x } + \alpha \boldsymbol { p }$
      \State $\boldsymbol { r } \leftarrow \boldsymbol { r } - \alpha \boldsymbol { s }$
      \State $\kappa \leftarrow \langle \boldsymbol { r } , \boldsymbol { r } \rangle$
      \If{$ \kappa < \epsilon$} \Comment{alt. fixed amount of steps}
        \State \Return \label{alg:conj_grad:exit}
      \EndIf
      \State $\beta \gets \frac{\kappa}{\gamma}$
      \State $\boldsymbol { p } \leftarrow \boldsymbol { r } + \beta \boldsymbol { p }$
      \State $\gamma \gets \kappa$
      \Until{false} \Comment{exit at \ref{alg:conj_grad:exit}}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}
\subsection{Framework}
decorator pattern
\subsection{Parallel Reduction}
\cite{NVIDIA:PARALLEL_REDUCTION}
\subsection{Binning}
\cite{NVIDIA:BINNING}
\subsection{Scanning}
\cite{NVIDIA:SCAN}
\cite{NVIDIA:SCAN_MODERN}
\subsection{Affine particle-in-cell}
\cite{MPM:APIC}
\subsection{Iterative solvers for linear systems}
%GMRES, Conjugate Gradients
\subsection{Verifying Simulation}
\begin{itemize}
  \item Conservation equations
    \cite{MPM:APIC}
  \item Singular Value Decomposition
  \item Parallel Reduction
\end{itemize}
\section{Evaluation}
\cite{NVIDIA:PEAK_PERFORMANCE}
\cite{KHRONOS:TIMER_QUERY}
\section{Conclusion}
\subsection{Review}
\subsection{Future Work}
Particle Activation: \cite{MPM:GPU}

>>>>

  \begin{equation}\label{eq:nodal_force}

@@

  \boldsymbol{h}(\boldsymbol{\hat{v}}_i^{n+1}) = m_i^n \frac{\boldsymbol{\hat{v}_i}^{n+1}-\boldsymbol{v}_i^n}{\Delta t} - \boldsymbol{f}_i\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{\hat{v}}_i^{n+1} + \boldsymbol{v}_i^{n})\right) = 0.

@@

In general starting off with an energy description is a more physical approach of the problem. The first term can be identified clearly as the kinetic energy. Moreover it easily allows adding potential terms for gravity or collision \cite{MPM:OPTIMI_INTEGR}. Or modifying the energy function altogether based off for instance the fundamental state of matter: as an example the phase transition to and physics of a liquid \cite{MPM:PHASE_CHANGE}.
The analysis of minimizing the objective is equivalent to finding the zero crossing of the derivative:
  \argmin_{\forall \boldsymbol{v}_i}(E(\boldsymbol{v}_i)) \Leftrightarrow g(\boldsymbol{v}_i) = \frac{\partial E}{\partial \boldsymbol{v}_i} = 0.
Since the objective from the start was minimizing $\boldsymbol{E}$ to find $\boldsymbol{v}_i$ the scale of this function can be chosen arbitrarily, i.e. a zero crossing does not scale. The scaling here can be identified as:
  \boldsymbol{g}(\boldsymbol{v}_i) = \frac{\Delta t}{4} h(\boldsymbol{v}_i) = m_i^n \frac{\boldsymbol{v_i}-\boldsymbol{v}_i^n}{4} -\frac{\Delta t}{4} \boldsymbol{f}_i\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}_i + \boldsymbol{v}_i^{n})\right).
\subsection{Newton's Method}
\newcommand{\var}[1]{{\ttfamily#1}}% variable
\begin{algorithm}[t]
  \caption{Conjugate gradient}\label{alg:conj_grad}
  \begin{algorithmic}[1]
    \Procedure{CONJUGATE-GRADIENT}{$H,x,f$}
    \State $\boldsymbol{x} \gets \Call{InitialGuess}$
    \State $\boldsymbol{Hx} \gets \Call{ComputeHp}{x}$
    \State $\boldsymbol{r}\gets \boldsymbol{f} - \boldsymbol{Hx}$
    \State $\boldsymbol{p}\gets \boldsymbol{r}$
    \State $\gamma \gets \langle \boldsymbol { r } , \boldsymbol { r } \rangle$
    \Repeat
      \State $\boldsymbol{Hp} \gets \Call{ComputeHp}{p}$
      \State $\boldsymbol { s } \gets \boldsymbol { H } \boldsymbol { p }$
      \State $\alpha \gets \frac { \gamma } { \langle \boldsymbol { p } , \boldsymbol { s } \rangle }$
      \State $\boldsymbol { x } \gets \boldsymbol { x } + \alpha \boldsymbol { p }$
      \State $\boldsymbol { r } \leftarrow \boldsymbol { r } - \alpha \boldsymbol { s }$
      \State $\kappa \leftarrow \langle \boldsymbol { r } , \boldsymbol { r } \rangle$
      \If{$ \kappa < \epsilon$} \Comment{alt. fixed amount of steps}
        \State \Return \label{alg:conj_grad:exit}
      \EndIf
      \State $\beta \gets \frac{\kappa}{\gamma}$
      \State $\boldsymbol { p } \leftarrow \boldsymbol { r } + \beta \boldsymbol { p }$
      \State $\gamma \gets \kappa$
      \Until{false} \Comment{exit at \ref{alg:conj_grad:exit}}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}
Many minimization and finding zero crossing algorithms are available. One of such is Newton's Method, which allows for rapid quadratic convergence in a near local neighborhood.
  \boldsymbol{v}_i^{(i+1)} = \boldsymbol{v}^{(i)}_i + \left[\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}\left(\boldsymbol{v}^{(i)}_i\right)\right]^{-1}\boldsymbol{g}\left(\boldsymbol{v}^{(i)}_i\right)
Computing the inverse is numerically irresponsible. Instead the following linear system is solved $\left(\Delta \boldsymbol{v} = \boldsymbol{v}_i^{(i+1)} - \boldsymbol{v}^{(i)}_i\right)$:
\begin{equation}\label{eq:linear_system}
  \left[\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}\left(\boldsymbol{v}^{(i)}_i\right)\right] \Delta \boldsymbol{v} =  \boldsymbol{g}\left(\boldsymbol{v}^{(i)}_i\right)
Using the Newton's Method however requires a computation of the Hessian of $E(\boldsymbol{v}_i)$:
  \frac{\partial \boldsymbol{g}_i}{\partial \boldsymbol{v}_j} = \frac{m_i^n}{4} - \frac{\Delta t^2}{16}\frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j}\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}_i + \boldsymbol{v}_i^{n})\right).
Computing $\frac{\partial f_i}{\partial x_j}$ for every combination $ij$ would be quite memory-intensive. Instead immediately the matrix-vector product on an increment $\delta \boldsymbol{u}_j$ is solved:
\begin{equation}\label{eq:hessian-matrix-vector}
  \sum_j\frac{\partial \boldsymbol{g}_i}{\partial \boldsymbol{v}_j}\delta \boldsymbol{u}_j = \frac{m_i^n}{4}\delta \boldsymbol{u}_j - \frac{\Delta t^2}{16}\sum_j \frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j}\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}_i + \boldsymbol{v}_i^{n})\right) \delta \boldsymbol{u}_j.
The linear system can be solved with the conjugate gradient method (Algorithm \ref{alg:conj_grad}) since $\frac{\partial g}{\partial v}$ is symmetric, positive definite. This is due to being the Hessian of the convex function $E$. There is to important things notice:
  \item Conservation of momentum on incomplete convergence is only fulfilled if \ref{eq:linear_system} is premultiplied by the inverse diagonal mass matrix $M^{-1}$. More in \cite{MPM:APIC}. Thus the equation becomes:
      \left(\frac{1}{4} - \frac{\Delta t^2}{16}\frac{1}{m_i^n}\frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j}\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}^{(i)}_i + \boldsymbol{v}_i^{n})\right)\right) \Delta v
      = \frac{\boldsymbol{v}_i^{(i)}-\boldsymbol{v}_i^n}{4} -\frac{\Delta t}{4}\frac{1}{m_i^n}\boldsymbol{f}_i\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}^{(i)}_i + \boldsymbol{v}_i^{n})\right).
  \item Due to \ref{eq:hessian-matrix-vector} the product of the Hessian with an increment ($\sum_j \frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j} \delta \boldsymbol{u}_j$) needs to be computed anew every step, since a pure matrix store is too costly.
The Hessian increment can be computed with a two-stage process:
  \item Compute a particle quantity $\boldsymbol{A}_p$ with a grid to particle transfer:
    \boldsymbol{A}_p = \frac { \partial ^ { 2 } \Psi } { \partial \boldsymbol { F } \partial \boldsymbol { F } } ( \hat { \boldsymbol { F } } _ { p } ( \hat { \boldsymbol { x } } )) : \left(\sum_j \delta \boldsymbol{u}_j (\nabla w_{jp}^n)^T \boldsymbol{F}^n_p\right).
  \item Compute the Hessian increment ($\delta \boldsymbol{f}_i$) with a particle to grid transfer:
  - \delta \boldsymbol{f}_i = - \sum_j \frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j} \delta \boldsymbol{u}_j = \sum_p V_p^0 \boldsymbol{A}_p(\boldsymbol{F}^n_p)^T \nabla w_{ip}^n.
\subsection{GPU memory}
\subsubsection{Optimization opportunities}
\cite{NVIDIA:BEST:PRACTICE}
\cite{AMD:GPU_OPEN}
\begin{itemize}
  \item Minimize CPU-GPU transfers
  \item Coalesced memory access
  \item Bank conflicts
\subsubsection{SoA vs. AoS}
\section{Implementation}
\subsection{Framework}
decorator pattern
\subsection{Parallel Reduction}
\cite{NVIDIA:PARALLEL_REDUCTION}
\subsection{Binning}
\cite{NVIDIA:BINNING}
\subsection{Scanning}
\cite{NVIDIA:SCAN}
\cite{NVIDIA:SCAN_MODERN}
\subsection{Affine particle-in-cell}
\subsection{Iterative solvers for linear systems}
%GMRES, Conjugate Gradients
\subsection{Verifying Simulation}
\begin{itemize}
  \item Conservation equations
  \item Singular Value Decomposition
  \item Parallel Reduction
\section{Evaluation}
\cite{NVIDIA:PEAK_PERFORMANCE}
\cite{KHRONOS:TIMER_QUERY}
\section{Conclusion}
\subsection{Review}
\subsection{Future Work}
Particle Activation: \cite{MPM:GPU}
